\Huge \bf \sc Implementation of a quadrotor UAV

### 2012-08-06 latex/paper/paper.tex
un poco + de paper. falta cosas de las figuras, la parte del vuelo y conc

% Meto paquetes a huevo
\usepackage{graphicx}

@@

A quadrotor platform can be imagined performing countless tasks. Many applications are being developed based on such platforms, but how do they fly? A aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining how to work with an IMU\cite{bib:calib_imu}\cite{bib:bola}\cite{bib:alain}, the development of the mathematical model \cite{bib:quadrotor-bible}\cite{bib:curso_quad} that represents the system, the filtering techniques applied to sensor data\cite{bib:kalman}\cite{bib:kalman2}, and the control system\cite{bib:quadrotor-bible}\cite{bib:hakas} that allows the quadrotor to fly.

The platform is based on a commercial radio controlled quadrotor, the frame and the motor control system were preserved, the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations requiered to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF, overcoming the problems inherent to each sensor and the noise generated by the vibrations of the motors, and providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm\cite{bib:lqrnotes}\cite{bib:lqr-discreto} is used to derive the control actions required to bring the system to the desired setpoint.

@@

\section{Model of a quadrotor}

A diagram of the quadrotor is shown in figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate clockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference are constantly used through out this paper: An intertial frame $S_I$ (relative to the Earth) $\lbrace \hat{i},\hat{j},\hat{k}\rbrace - \lbrace \text{North},\text{West},\text{Up}\rbrace$) and a non-intertial frame $S_q$ relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations show in figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are know as Euler angles.

\begin{figure} [h!]
  \centering
  \subfloat[Rotation 1: Axis $\hat{k}$]{\label{fig:angulos_1}
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_1.png}}
  \subfloat[Rotation 2: Axis $\hat{j}$]{\label{fig:angulos_2} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_2.png}}
  \subfloat[Rotation 3: Axis $\hat{i}$]{\label{fig:angulos_3} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_3.png}}
  \caption{Rotations applied on $S_I$ to obtain $S_q$.}
  \label{fig:rotaciones}
\end{figure}

From a detailed analysis of the dynamics and kinematics of the quadrotor, the state vector $\vec{X}$ given by (\refp{eq:sv}) is built to describe the system at any given time. Practical considerations lead to choosing the state variables referenced to the quadrotor frame $S_q$. This choice introduces simplifications in the theoretical development, and simplifies the interpretation of the data provided by the IMU, which is mounted on the quadrotor and hence provides provides accelerations and angular velocities that are relative to $S_q$:
\begin{equation}
\vec{X}=\left\lbrace  x,y,z, \theta,\varphi,\psi, v_{q_z},v_{q_y},v_{q_z},\omega_{q_x},\omega_{q_y},\omega_{q_z} \right\rbrace
\label{eq:sv}
\end{equation}
where:
\begin{itemize}
\item $\lbrace x,y,z \rbrace$ are the cartesian coordinates in $S_I$.
\item $\lbrace\theta,\varphi,\psi\rbrace$ are the Euler angles show in (\refp{fig:rotaciones}).
\item $\lbrace v_{q_x},v_{q_y},v_{q_z}\rbrace$ are the linear velocities relative to $S_I$.
\item $\lbrace w_{q_x},w_{q_y},w_{q_z}\rbrace$ are the angular velocities relative to $S_q$ (right hand rule applied on $\lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$).
\end{itemize}
%where $\lbrace x,y,z \rbrace$ are the cartesian coordinates relative to the inertial frame $S_I$ (mapped on $\lbrace \hat{i},\hat{j},\hat{k}\rbrace$), $\lbrace\theta,\varphi,\psi\rbrace$ are the Euler angles show in (\refp{fig:rotaciones}), $\lbrace v_{q_x},v_{q_y},v_{q_z}\rbrace$ are the linear velocities relative to $S_I$ and $\lbrace w_{q_x},w_{q_y},w_{q_z}\rbrace$ are the angular velocities relative to $S_q$ (right hand rule applied on $\lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$).

The equations that govern the dynamics of the quadrotor, which are highly non-linear, are presented in equation \refp{eq:modelo}.
\begin{equation}
\small
\boxed{\begin{aligned}&\dot{x}\!=\!v_{q_x} \cos \varphi \cos \theta \!+\! v_{q_y} ( \cos \theta \sin \varphi \sin \psi\!-\!\cos \varphi \sin \theta )\\
&\quad \!+\! v_{q_z}(\sin \psi \sin \theta \!+\! \cos \psi \cos \theta \sin \varphi)\\
&\dot{y}\!=\!v_{q_x} \cos \varphi \sin \theta \!+\! v_{q_y} (\cos \psi \cos \theta \!+\! \sin \theta \sin \varphi \sin \psi)\\
&\quad\!+\! v_{q_z}( \cos \psi \sin \theta \sin \varphi\!-\!\cos \theta \sin \psi )\\
&\dot{z}\!=\! \!-\!v_{q_x} \sin \varphi  \!+\! v_{q_y} \cos \varphi \sin \psi  \!+\! v_{q_z}\cos \varphi \cos \psi\\
&\dot{\psi}\!=\!\omega_{q_x} \!+\! \omega_{q_z}\tan\varphi \cos\psi \!+\! \omega_{q_y}\tan\varphi \sin\psi\\
&\dot{\varphi}\!=\!\omega_{q_y}\cos \psi \!-\! \omega_{q_z}\sin\psi\\
&\dot{\theta}\!=\!\omega_{q_z} \frac{\cos\psi}{\cos\varphi}  \!+\! \omega_{q_y}\frac{\sin\psi}{\cos\varphi}\\
&\dot{v_{q_x}}\!=\!v_{q_y} \omega_{q_z} \!-\! v_{q_z} \omega_{q_y}\!+\!g\sin\varphi\\
&\dot{v_{q_y}}\!=\!v_{q_z} \omega_{q_x} \!-\! v_{q_x} \omega_{q_z}\!-\!g\cos\varphi\sin\psi\\
&\dot{v_{q_z}}\!=\!v_{q_x} \omega_{q_y} \!-\! v_{q_y} \omega_{q_x}\!-\!g\cos\varphi\cos\psi\!+\!\frac{1}{M}\sum_{i\!=\!1}^4T_i\\
&\dot{\omega_{q_x}}\!=\!\frac{1}{I_{xx}}\omega_{q_y}\omega_{q_z}(I_{yy}\!-\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{xx}}\omega_{q_y}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{xx}}dMg\cos\varphi\sin\psi\!+\!\frac{1}{I_{xx}}L(T_2\!-\!T_4)\\
&\dot{\omega_{q_y}}\!=\!\frac{1}{I_{yy}}\omega_{q_x}\omega_{q_z}(\!-\!I_{xx}\!+\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{yy}}\omega_{q_x}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{yy}}dMg\sin\varphi\!+\!\frac{1}{I_{yy}}L(T_3\!-\!T_1)\\
&\dot{\omega_{q_z}}\!=\!\frac{1}{I_{zz}}\left(\!-\!Q_1\!+\!Q_2\!-\!Q_3\!+\!Q_4\right)\\
%\frac{\!-\!I_{zz_m}(\dot{w_1}\!-\!\dot{w_2}\!+\!\dot{w_3}\!-\!\dot{w_4})}{I_{zz}}\!+\!
\end{aligned}}
\label{eq:modelo}
\end{equation}
To simplifly the control system, equations \refp{eq:modelo} are linearized near certain points of operation which result in a Linear Time Invariant (LTI) system. These points allow the following trajectories: Hovering\footnote{Constant state vector.}, uniform linear trajectories, and uniform circular trajectories. The system will be restricted to these three trajectories, since it greatly simplifies the control system and does it not introduce significant limitations.

%integradores -> permiten tomar en consideracion cosas q le cuelges, desequilibrios

@@

@@

Every sensor has its issues: The gyro drifts over time; the accelerometer is very sensible to the vibrations generated by the motors; the magnetometer is distorted by ferromagnetic materials; the GPS has very little precision and a slow update rate. Each sensor by itself is very limited, but they can be combined to compensate for their limitations. The filter takes care of this.
%TODO pasar figura a ingl\'e
\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{../pics_kalman/diagrama_kalman_2.png}
	\caption{\textbf{EKF - } Outline of how sensor data is combined to estimate the state variables.}
	\label{fig:diagrama_kalman}
\end{figure}

@@

@@

>>>>

% Meto paquetes a huevo
\usepackage{graphicx}
\usepackage[table]{xcolor}

@@

A quadrotor platform can be imagined performing countless tasks. Many applications are being developed based on such platforms, but how do they fly? An aerial vehicle is inherently unstable, staying still is not a simple task. This paper focuses on the stabilization of a quadrotor, explaining the development of the mathematical model \cite{bib:quadrotor-bible}\cite{bib:curso_quad} to represent the system, the filtering techniques applied for sensor data fusion\cite{bib:kalman}\cite{bib:kalman2} and the control system\cite{bib:quadrotor-bible}\cite{bib:hakas} that allows the quadrotor to fly.

The platform is based on a commercial radio controlled quadrotor, the frame and the motor control system were preserved, the IMU and intelligence were replaced by the flight controller that was developed. A BeagleBoard\footnote{BeagleBoard development board - \url{http://beagleboard.org/}} running Linux\footnote{Angstrom distribution: \url{http://www.angstrom-distribution.org/}} performs the computations requiered to convert raw data received from the IMU\footnote{\textit{Mongoose} IMU - \url{http://store.ckdevices.com/}} over a UART and combine it using an EKF. The EKF overcomes the problems inherent to each sensor and filters out noise, providing a reliable estimation of the state vector. Once the current state is known, the LQR algorithm\cite{bib:lqrnotes}\cite{bib:lqr-discreto} is used to derive the control actions required to bring the system to the desired setpoint.

@@

\section{Model of a quadrotor}
\label{sec:modelo}

\subsection{Definitions}
\label{sec:modelo-defs}

A diagram of the quadrotor is shown in figure \refp{fig:quad}. Two of the motors rotate clockwise (2 and 4) and the other two (1 and 3) rotate clockwise. This configuration allows the quadrotor to rotate, tilt and gain/lose altitud by setting different speeds on each motor. Two frames of reference (figure \ref{fig:quad}) are constantly used through out this paper: An intertial frame $S_I$, relative to the Earth, and a non-intertial frame $S_q$, relative to the quadrotor. The mapping of one frame to the other can be achieved by applying the three rotations shown in figure \refp{fig:rotaciones}. The angles $\lbrace \theta, \varphi, \psi\rbrace$ are known as Euler angles.

\begin{figure} [h!]
  \centering
  \subfloat[Rotation 1: Axis $\hat{k}$]{\label{fig:angulos_1}
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_1.png}}
  \subfloat[Rotation 2: Axis $\hat{j}$]{\label{fig:angulos_2} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_2.png}}
  \subfloat[Rotation 3: Axis $\hat{i}$]{\label{fig:angulos_3} 
  		\includegraphics[width=0.3\columnwidth]
  			{../pics_modelo_fisico/angulos_3.png}}
  \caption{\textbf{Mapping:} Rotations applied on $S_I$ to obtain $S_q$.}
  \label{fig:rotaciones}
\end{figure}

\subsection{Dynamics-kinamatics of the system}
\label{sec:modelo-dyn-kin}

From a detailed analysis of the dynamics and kinematics of the quadrotor, the equations \refp{eq:modelo} are obtained, and the state vector shown in \refp{eq:sv} is built to describe the system at any given time. The variables with subscript \textit{q} are referenced to the quadrotor frame $S_q$, the rest are relative to $S_i$. This choice simplifies the theoretical development and the interpretation of the data provided by the IMU, which is mounted on the quadrotor and hence provides provides accelerations and angular velocities that are relative to $S_q$:
\begin{equation}
\vec{X}=\left\lbrace  x,y,z, \theta,\varphi,\psi, v_{q_z},v_{q_y},v_{q_z},\omega_{q_x},\omega_{q_y},\omega_{q_z} \right\rbrace
\label{eq:sv}
\end{equation}
where:
\begin{itemize}
\item $\lbrace x,y,z \rbrace$ are the cartesian coordinates in $S_I$.
\item $\lbrace\theta,\varphi,\psi\rbrace$ are the Euler angles show in (\refp{fig:rotaciones}).
\item $\lbrace v_{q_x},v_{q_y},v_{q_z}\rbrace$ are the linear velocities relative to $S_I$.
\item $\lbrace w_{q_x},w_{q_y},w_{q_z}\rbrace$ are the angular velocities relative to $S_q$ (right hand rule applied on $\lbrace \hat{i}_q,\hat{j}_q,\hat{k}_q\rbrace$).
\end{itemize}

\begin{equation}
\small
\boxed{\begin{aligned}&\dot{x}\!=\!v_{q_x} \cos \varphi \cos \theta \!+\! v_{q_y} ( \cos \theta \sin \varphi \sin \psi\!-\!\cos \varphi \sin \theta )\\
&\quad \!+\! v_{q_z}(\sin \psi \sin \theta \!+\! \cos \psi \cos \theta \sin \varphi)\\
&\dot{y}\!=\!v_{q_x} \cos \varphi \sin \theta \!+\! v_{q_y} (\cos \psi \cos \theta \!+\! \sin \theta \sin \varphi \sin \psi)\\
&\quad\!+\! v_{q_z}( \cos \psi \sin \theta \sin \varphi\!-\!\cos \theta \sin \psi )\\
&\dot{z}\!=\! \!-\!v_{q_x} \sin \varphi  \!+\! v_{q_y} \cos \varphi \sin \psi  \!+\! v_{q_z}\cos \varphi \cos \psi\\
&\dot{\psi}\!=\!\omega_{q_x} \!+\! \omega_{q_z}\tan\varphi \cos\psi \!+\! \omega_{q_y}\tan\varphi \sin\psi\\
&\dot{\varphi}\!=\!\omega_{q_y}\cos \psi \!-\! \omega_{q_z}\sin\psi\\
&\dot{\theta}\!=\!\omega_{q_z} \frac{\cos\psi}{\cos\varphi}  \!+\! \omega_{q_y}\frac{\sin\psi}{\cos\varphi}\\
&\dot{v_{q_x}}\!=\!v_{q_y} \omega_{q_z} \!-\! v_{q_z} \omega_{q_y}\!+\!g\sin\varphi\\
&\dot{v_{q_y}}\!=\!v_{q_z} \omega_{q_x} \!-\! v_{q_x} \omega_{q_z}\!-\!g\cos\varphi\sin\psi\\
&\dot{v_{q_z}}\!=\!v_{q_x} \omega_{q_y} \!-\! v_{q_y} \omega_{q_x}\!-\!g\cos\varphi\cos\psi\!+\!\frac{1}{M}\sum_{i\!=\!1}^4T_i\\
&\dot{\omega_{q_x}}\!=\!\frac{1}{I_{xx}}\omega_{q_y}\omega_{q_z}(I_{yy}\!-\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{xx}}\omega_{q_y}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{xx}}dMg\cos\varphi\sin\psi\!+\!\frac{1}{I_{xx}}L(T_2\!-\!T_4)\\
&\dot{\omega_{q_y}}\!=\!\frac{1}{I_{yy}}\omega_{q_x}\omega_{q_z}(\!-\!I_{xx}\!+\!I_{zz})\\
&\quad\!+\!\frac{1}{I_{yy}}\omega_{q_x}I_{zz_m}(\omega_1\!-\!\omega_2\!+\!\omega_3\!-\!\omega_4)\\
&\quad\!-\!\frac{1}{I_{yy}}dMg\sin\varphi\!+\!\frac{1}{I_{yy}}L(T_3\!-\!T_1)\\
&\dot{\omega_{q_z}}\!=\!\frac{1}{I_{zz}}\left(\!-\!Q_1\!+\!Q_2\!-\!Q_3\!+\!Q_4\right)\\
%\frac{\!-\!I_{zz_m}(\dot{w_1}\!-\!\dot{w_2}\+!\!\dot{w_3}\!-\!\dot{w_4})}{I_{zz}}\!+\!
\end{aligned}}
\label{eq:modelo}
\end{equation}

@@

In order to determine what actions should be taken, the state of the system must be known. The system uses a 9 degrees of freedom IMU and a GPS. This equipment enables direct measurement of most of the state variables. There is no direct measurement of the linear speed of the system $\{v_{q_z},v_{q_y},v_{q_z}\}$, so the model developed in \ref{sec:modelo} is used to estimate them.

\subsection{IMU}
\label{sec:sensors-imu}

The IMU is equiped with the following sensors:
\begin{itemize}
\item \textbf{Barometer:} Measures the absolute pressure of the environment. Variations of pressure are used to estimate variations in the altitud of the system.
\item \textbf{Thermometer:} The barometer includes a thermometer, which is used internally by the barometer. The temperature data is used to apply a temperature compensation to the calibrations performed on the gyroscope and the accelerometer.
\item \textbf{Gyroscope:} A 3-axis gyroscope is used to measure angular velocity. A calibration \cite{bib:calib_imu} was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Accelerometer:} A 3-axis accelerometer is used to measure gravity. Under the hypothesis that no other accelerations are present, it is used to determine two of the three Euler angles: $\{\psi, \phi\}$. This hypothesis is acceptalbe, since the rest of the accelerations involved are not significant compared to gravity. A calibration \cite{bib:calib_imu} was designed and applied to this device, as well as a temperature-compensation.
\item \textbf{Magnetometer:} In an area free of magnetic interference this 3-axis sensor will measure the Earth's magnetic field, allowing to determine what direction is North. If the system is horizontal (or the inclination is estimated using other sensors) this sensor can be used to determine the last of the three Euler angles: $\theta$. A calibration was performed on this sensor, based on \cite{bib:bola} and \cite{bib:alain}.
\end{itemize}

\subsection{GPS}
\label{sec:sensors-gps}

A GPS is used to determine the absolute position of the system, since the IMU cannot provide this information. The accuracy under good sky visibility is of 2-3 meters. The GPS's performance improves when the system is moving.

\subsection{Sensor Specifications}
\label{sec:sensors-specifications}

Table \ref{tab:sensors:resumen} shows an outline of the specifications of the sensors used.

\begin{table}[h]
\begin{center}
\rowcolors{1}{gray!20}{}
\begin{tabular}{|c|c|c|}
\hline
 & Rate & Resolution \\
\hline
Accelerometer XY & 10ms (x2)& 4mg\\
\hline
Accelerometer Z  & 10ms (x2)& 4mg\\
\hline
Gyro XY  & 10ms (x2)& 0.07 $^o/s$\\
\hline
Gyro Z  & 10ms (x2)& 0.07 $^o/s$\\
\hline
Barometer  & 10ms (x1) & 1Pa\\
\hline
Magnetometer XY  & 10ms  (x2)& 5 mGa\\
\hline
Magnetometer Z  & 10ms (x2)& 5 mGa\\
\hline
GPS  & 1s & - \\
\hline
\end{tabular}
\end{center}
\caption{\textbf{Sensor specifications:} A rate of ``\textit{10ms (x2)}'' means that every 10ms the result of averaging 2 samples is received from the IMU.}
\label{tab:sensors:resumen}
\end{table}

@@

Every sensor has its issues: The gyro drifts over time; the accelerometer is very sensible to the vibrations generated by the motors; the magnetometer is distorted by ferromagnetic materials; the GPS has very little precision and a slow update rate. Each sensor by itself is very limited, but they can be combined to compensate for their limitations. The filter takes care of this.

%TODO pasar figura a ingl\'e
\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{../pics_kalman/diagrama_kalman_2.png}
	\caption{\textbf{EKF - } Outline of how sensor data is combined to estimate the state variables.}
	\label{fig:diagrama_kalman}
\end{figure}

@@

Figure \refp{fig:diagrama_kalman} shows a diagram of how data from the sensors is combined within the EKF, assisted by the model of the system. The output of the filter is an estimation the all the variables of the state vector.

@@

To work directly with equations \refp{eq:modelo}, a non-linear control technique would be required. To simplifly the control system, equations \refp{eq:modelo} are linearized near certain points of operation which result in a Linear Time Invariant (LTI) system. This restricts the system to: Hovering\footnote{Constant state vector.}, uniform linear trajectories, and uniform circular trajectories. This choice greatly simplifies the control system and does it not introduce significant limitations.

The control system is shown in figure \ref{fig:diagrama_bloques_eng.pdf}. The feedback matrices have many entries, $K_p$ has 48 and $K_i$ has 16. A \textit{root locus} or \textit{pole-zero} analysis on such a system is not a simple task, additional complications are introduced by the fact that $K_p$ and $K_i$ will change when the trayectory changes, they depend on the linearization of the system. A simple and automated solution is achieved by means of the LQR algorithm. The initial design did not include the $K_i$ term, it was introduced to take into consideration errors in the characterization and modelling of the system.

\begin{figure}
	\centering
	\includegraphics[width=1\columnwidth]{../pics_control/diagrama_bloques_eng.pdf}
\vspace{-220pt}
	\caption{\textbf{Control System:} $x_{sp}$ and $u_{sp}$ represent the setpoints for the state vector and the speed of each motor. The $K_p$ and $K_i$ blocks are the proportional and integral gain matrices. The output of the system is the current state vector.}
	\label{fig:diagrama_bloques_eng.pdf}
\end{figure}

\subsection{The LQR algorithm}
\label{sec:lqr}

For a system which is controllable and observable\cite{bib:hakas}, the LQR algorithm finds an optimal controller\cite{bib:lqrnotes} by minimizing the following cost function:

\begin{equation}
\label{eq:lqr}
\int_{0}^{\infty}  \vec{X}'(t)Q \vec{X}(t)+\vec{U}'(t)R\vec{U}(t)dt
\end{equation}

where $Q$ and $R$ are positive-definite matrices, representing the energy of the controlled states and the control signal, respectively. The choice of $Q$ and $R$ is a tradeoff between the error that is tolerated and the energy that the system can use to correct an error.

 When the system is a feedback system with known state variables, like the one shown in figure \refp{fig:diagrama_bloques_eng.pdf}, the solution is given by \refp{eq:lqr-sol}, where $P$ is the solution to Ricatti's equation.

\begin{equation}
\begin{array}{c}
\vec{U}(t)-\vec{U}^*(t) = -K(\vec{X}(t)-\vec{X}^*(t))\\
\\
K = R^{-1}B^TP
\end{array}
\label{eq:lqr-sol}
\end{equation}

For the controllability and observability hypothesis to hold, only some state variables may be passed through the integral control block: $\{x,y,z,\theta\}$.
